%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%12pt: grandezza carattere
%a4paper: formato a4
%openright: apre i capitoli a destra
%twoside: serve per fare un documento fronteretro
%report: stile tesi (oppure book)
\documentclass[12pt,a4paper,openright,twoside]{report}

\usepackage{tabularx}
\usepackage[english,provide=*]{babel}
\usepackage[utf8]{inputenc}
\usepackage{fancyhdr}
\usepackage{indentfirst}
\usepackage{graphicx}
\usepackage{newlfont}
%librerie matematiche
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{latexsym}
\usepackage{amsthm}
\usepackage{listings}
\usepackage{chngcntr}
\usepackage[outputdir=build,chapter]{minted}
\usepackage{algpseudocodex}

\oddsidemargin=30pt \evensidemargin=20pt%impostano i margini
\hyphenation{sil-la-ba-zio-ne pa-ren-te-si}

\usepackage[square, numbers, comma, sort&compress]{natbib} %Bibliografia
\usepackage[hidelinks]{hyperref}
\usepackage{float}
\usepackage{caption}
\usepackage[pass]{geometry}

%comandi per l'impostazione della pagina, vedi il manuale della libreria fancyhdr per ulteriori delucidazioni
\captionsetup{font=small}
\setlength{\parindent}{0pt}
\pagestyle{fancy}\addtolength{\headwidth}{20pt}
\renewcommand{\chaptermark}[1]{\markboth{\thechapter.\ #1}{}}
\renewcommand{\sectionmark}[1]{\markright{\thesection \ #1}{}}
\rhead[\fancyplain{}{\bfseries\leftmark}]{\fancyplain{}{\bfseries\thepage}}
\cfoot{}

\linespread{1.3} %comando per impostare l'interlinea

\renewcommand{\listingscaption}{Codice}
\renewcommand{\listoflistingscaption}{List of Algorithms}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% FRONTESPIZIO
\newgeometry{hmarginratio=1:1}
\input{frontespizio.tex}
\restoregeometry

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT
\clearpage{\pagestyle{empty}\cleardoublepage}
\pagenumbering{roman}
\renewcommand{\abstractname}{Introduction}
\phantomsection
\addcontentsline{toc}{chapter}{Introduction}
\begin{abstract}
  In recent years, advancements in distributed computing and decentralized technologies have reshaped the
  landscape of machine learning and data processing. This document explores the intersection of two
  innovative paradigms: Federated Learning and Blockchain. Each of these fields has gathered
  significant attention for its unique potential — federated learning for enabling collaborative model training
  without centralized data aggregation, and blockchain for providing decentralized governance and immutable ledgers
  with robust security guarantees.
  
  Federated learning is a decentralized approach to training machine learning models collaboratively across
  multiple devices or nodes while keeping data local to each participant. Originally developed to address
  privacy concerns, Federated Learning has found applications in domains such as healthcare, mobile devices, and IoT, where
  sensitive data remains confined to the edge devices that generate it. However, Federated Learning is not without its
  challenges.
  
  Blockchain, on the other hand, is a decentralized ledger technology that ensures secure and immutable
  decentralized data storage and trustless interactions among participants.
  First popularized by cryptocurrencies such as Bitcoin, blockchain has since evolved to support a wide range
  of use cases, including supply chain management, smart
  contracts, and secure data sharing.
  
  The possibility of integrating blockchain with federated learning has sparked interest in the research community as a
  means to address some of the inherent limitations of the paradigm.
  Federated learning's reliance on a central server for model aggregation introduces risks of a single point of failure
  and trust issues among participants.
  Blockchain's decentralized nature and consensus mechanisms can solve this problem among others, such as
  ensuring model integrity and fairness in participation. However this comes with its own set of challenges,
  mostly communication, storage and (depending on the underlying mechanism of the blockchain, also computation) overhead.
  
  Integrating blockchain into federated learning systems presents unique challenges. These include the
  computational and communication overhead introduced by blockchain protocols, ensuring scalability in systems
  with millions of devices, and aligning the asynchronous nature of federated learning with blockchain's
  consensus requirements. Despite these hurdles, the potential benefits—such as enhanced security,
  transparency, and trust—make this integration a promising avenue of research.
  
  This thesis aims to explore the foundational concepts of both federated learning and blockchain, 
  analyze their potential integration, and address the challenges that arise from combining these 
  technologies. Through theoretical insights and practical implementation, the study will evaluate the 
  feasibility and benefits of using blockchain to enhance federated learning systems.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% RINGRAZIAMENTI
% \clearpage{\pagestyle{empty}\cleardoublepage}
% \chapter*{Ringraziamenti}
% \phantomsection
% \addcontentsline{toc}{chapter}{Ringraziamenti}
% TODO

\clearpage{\pagestyle{empty}\cleardoublepage}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% INDICE
\tableofcontents %crea l'indice
%imposta l'intestazione di pagina
\rhead[\fancyplain{}{\bfseries\leftmark}]{\fancyplain{}{\bfseries\thepage}}
\lhead[\fancyplain{}{\bfseries\thepage}]{\fancyplain{}{\bfseries
Index}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ELENCO DELLE FIGURE
\clearpage{\pagestyle{empty}\cleardoublepage}
\renewcommand{\listfigurename}{List of Figures}
\phantomsection
\addcontentsline{toc}{chapter}{List of Figures}
\listoffigures %crea l'elenco delle figure

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ELENCO DELLE TABELLE
\clearpage{\pagestyle{empty}\cleardoublepage}
\renewcommand{\listtablename}{List of Tables}
\phantomsection
\addcontentsline{toc}{chapter}{List of Tables}
\listoftables %crea l'elenco delle tabelle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ELENCO DEI CODICI
\clearpage{\pagestyle{empty}\cleardoublepage}
\phantomsection
\addcontentsline{toc}{chapter}{List of Algorithms}
\listoflistings %crea l'elenco dei codici

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% INTRODUZIONE
\clearpage{\pagestyle{empty}\cleardoublepage}
\markboth{}{} % Clears the header markers
\pagenumbering{arabic} %mette i numeri arabi
\lhead[\fancyplain{}{\bfseries\thepage}]{\fancyplain{}{\bfseries\rightmark}}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% SCENARI APPLICATIVI E STATO DELL'ARTE

\clearpage{\pagestyle{empty}}
\include{chapters/blockchain/index}

\clearpage{\pagestyle{empty}}
\include{chapters/federated-learning/index}

\clearpage{\pagestyle{empty}}
\include{chapters/integration/index}

\clearpage{\pagestyle{empty}}
\include{chapters/conclusion/index}

% BIBLIOGRAFIA
\nocite{*}
\addcontentsline{toc}{chapter}{Bibliography}
\label{Bibliography}
\bibliographystyle{IEEEtran}
\bibliography{bibliography}
\rhead[\fancyplain{}{\bfseries Bibliography}]
{\fancyplain{}{\bfseries\thepage}}
\end{document}
