\chapter{Federated Learning and Machine Learning} \label{chap:federated-learning}
\section{Introduction to machine learning}
One of the central topics of this document is Federated Learning, however, before introducing it in depth, we
must first introduce the subject to which it belongs: machine learning.
What is machine learning and what is its underlying problem? Machine
learning is the discipline which seeks to answer the following
question: \emph{"how can a computer improve based on experience?"}.

There are three main branches of machine learning:
\begin{enumerate}
  \item Supervised learning
  \item Unsupervised learning
  \item Reinforcement learning
\end{enumerate}

Federated learning was originally born as a supervised learning technique, however recent work has been done to bring
ito unsupervised \cite{UnsupervisedFL} and reinforcement learning \cite{ReinforcementFL}.
A short summary of all three learning methods is the following:
\begin{itemize}
  \item Supervised learning about generalizing a dataset of examples to produce a general rule that can be
    applied to new examples.
  \item Unsupervised learning is about finding patterns and relationship between examples of the dataset.
  \item Reinforcement Learning studies how well a machine is able to adapt to a dynamic environmnet based on
    reward and penalties.
    When the machine does 'well' it receives a reward, otherwise it recieves a penalty.
\end{itemize}
For the sake of this document we will only cover supervised learning in a little more depth.

If the reader is interested in diving deeper into the topic of machine learning and gain a more thorough
understanding of the techniques adopted and discussed here we suggest to take a look at \cite{MLBook}.

\input{chapters/federated-learning/supervised-learning.tex}
\clearpage

\input{chapters/federated-learning/gradient-descent.tex}
\clearpage

\section{Introduction to Federated Learning}
To \textit{federate} means "to join together under a central government or organization while keeping some
local control" \cite{oxford_federate}.
This principle forms the foundation of federated learning: uniting many independent devices under the
coordination of a central server to collaboratively train a global machine learning model, all while ensuring
that data remains decentralized and local to the devices.

Federated learning represents a significant paradigm shift in the field of machine learning, distinct from
both traditional and distributed approaches.
\textbf{Traditional machine learning} methods often rely on a centralized architecture, where all data is
collected and stored in a central server. This centralization simplifies the training process but raises
concerns related to data privacy, security, and scalability when dealing with sensitive or large-scale datasets.

\textbf{Distributed machine learning}, on the other hand, addresses scalability by distributing the
computational workload across multiple nodes or servers. These nodes process subsets of data stored centrally
or across a cluster. However, distributed learning still typically assumes that data can be transferred and
aggregated freely, making it less suited for scenarios where privacy or data sovereignty is a primary concern.

\textbf{Federated learning} bridges the gap between these approaches by introducing a decentralized training
methodology (with respect to the data). In federated learning, data remains on the devices where it is
generated (e.g., smartphones, IoT
devices, or edge servers), and only model updates—such as gradients or parameters—are transmitted to the
central server for aggregation. This design inherently enhances privacy and security, minimizes data
movement, and opens opportunities for leveraging massive amounts of distributed data that were previously
inaccessible under traditional paradigms.

In this chapter, we will explore the mechanisms and architectures underpinning federated learning, starting
with its foundational model, also called \textit{vanilla federated learning}, and explore alternative
approaches such as integrating it with blockchain techonology.

\input{chapters/federated-learning/vanilla-fl.tex}
